# Get map data {#sec-chap013}


```{r}
#| label: setup
#| results: hold
#| include: false

base::source(file = "R/helper.R")

```


::::: {#obj-chap013}
:::: {.my-objectives}
::: {.my-objectives-header}
Objectives
:::

::: {.my-objectives-container}

Download map data from `r glossary("Natural Earth")`.

:::
::::
:::::

With {**rnaturalearth**} there is a special package for downloading the geometries of all the countries in the world. I will take the medium scale ('50') of the 'physical' category. The data source is [Natural Earth](https://www.naturalearthdata.com/).

I downloaded original data from Natural Earth with `rnaturalearth::ne_download()`  and compared it with the pre-downloaded data from the `rnaturalearth::ne_countries()`. It turned out that they are essentially identical. So I used the pre-downloaded data with the advantage that the resulting files have not converted from  `.shp` files to a {**sf**} file.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-013-get-map-data}
: Get map data from Natural Earth
::::::
:::
::::{.my-r-code-container}

<center>**Run this code chunk manually if the file(s) still needs to be downloaded.**</center>

::: {#lst-013-get-map-data}

```{r}
#| label: get-map-data
#| eval: false

raw_countries <- rnaturalearth::ne_countries(
  scale = "medium",
  type = "countries",
  continent = NULL,
  country = NULL,
  geounit = NULL,
  sovereignty = NULL,
  returnclass = "sf"
)

my_save_data_file("chap013", raw_countries, "raw_countries.rds")

```

Get map data from Natural Earth
::: 
::::
:::::

## Glossary Entries {.unnumbered}

```{r}
#| label: glossary-table
#| echo: false

glossary_table()
```

------------------------------------------------------------------------

## Session Info {.unnumbered}

::: my-r-code
::: my-r-code-header
Session Info
:::

::: my-r-code-container
```{r}
#| label: session-info

sessioninfo::session_info()
```
:::
:::

